Taking Conditional Independence Seriously:\\A Latent Network Approach to Dyadic Analysis

### 2017-10-27 paper/netsMatter.tex
new(ish) abstract

\small{\singlespacing{
\begin{abstract}
	International relations scholarship is filled with dyads. This includes dyadic hypotheses and propositions, but especially data. These relational data contain information about the interdependencies of various phenomena, including countries, dyads, and even triads. However, most empirical studies of such data do not take into account these dependencies. As a result, the studies are often contradictory and produce findings that are not compelling. One reason is that the independence required by typical methods does not exist in the data. We present a different, regression based method, which constructs a latent network and incorporates \first, \second, and \third-order dependencies. We replicate five prominent studies in recent IR scholarship and compare the standard approach to the latent factor approach. The additive and multiplicative latent factor approach is shown to produce more precise estimates of covariate effects and it also dominates standard approaches in terms of out-of-sample cross-validations.
\end{abstract}
}}

>>>>

\small{\singlespacing{
\begin{abstract}
	%International relations scholarship is filled with dyads. This includes dyadic hypotheses and propositions, but especially dyadic data. Yet, much of the work conducted brushes aside the statistical complications that come with analyzing relational data, namely that the observations do not satisfy the conditional independence criterion. As a result, these studies often produce results with biased coefficient estimates and poorly calibrated standard errors, factors which have consequences for hypothesis testing. We present an alternative regression based approach that accounts for the dependencies that complicate inferential analysis in the context of relational data using a set of additive and multiplicative effects within a Bayesian framework. To highlight the utility of this framework for studying relational data, we present a simulation exercise highlighting the model's ability to account for the dependencies that emerge in relational data. Additionally, we replicate five prominent studies in recent IR scholarship and compare the standard approach to the latent factor approach. For each study, our approach shows that conventional methods overstate the effect of key variables in the study of interstate conflict, underestimate the uncertainty in these effects, and in some cases lead researchers to faulty conclusions about the statistical significance of their variables. Further, the new approach dominates in terms of out-of-sample cross-validations, rendering it more useful in forecasting applications and in modeling the data generating process behind outcomes of interest.

Much international relations scholarship concerns dyads: dyadic hypotheses and propositions, and especially dyadic data. Yet, much contemporary research brushes aside the complications that come with analyzing relational data that are interdependent. Dyadic observations do not typically satisfy the conditional independence criterion required for many statistical approaches. As a result, many studies often produce results with biased coefficient estimates and poorly calibrated standard errors. These biases have profound consequences for evaluating parametric models. 
%parahere?
We present an alternative, regression-based approach that accounts for the dependencies that complicate inferential analysis in the context of relational data. It uses a set of additive and multiplicative effects within a Bayesian regression framework. To highlight the utility of this framework for studying relational data, we present a simulation highlighting the ability of this approach to account for the dependencies that frequently characterize relational data such as often utilized in international relations. In addition, we replicate five prominent studies in recent international relations scholarship, comparing the standard approach to the proposed latent factor approach. For each study, we find that conventional methods overstate the effect of key variables in the study of interstate conflict (and trade), underestimate the uncertainty in these effects, and in some cases lead researchers to faulty conclusions about the statistical significance and substantive importance of their variables and the hypotheses they derive from. The latent factor approach also dominates the current status quo approach in terms of out-of-sample cross-validations, rendering it more useful in hypothesis testing as well as forecasting. In short, it seems able to better capture the data generating process behind outcomes of interest.
%parahere?
We summarize the features of this latent factor model, and provide guideance on how to implement it in additional studies.
\end{abstract}
}}
